---
output:
  pdf_document: default
  html_document: default
---
# Chapter 1: Installation and set up {#ch1}

## Installing R and RStudio 

Before we can get started with network analysis, we need to prepare a computing environment for doing so. We will analyze networks using the `igraph` R package, and so the very first step is to install both R and RStudio into the device that you wish to use for network analysis. It is important to note that R and RStudio are two different things that you have to install. R is the programming language itself (like Python or C++) whereas RStudio is the software application that we will use to interact with R. RStudio is known as an *integrated development environment* (other examples of IDEs are Visual Studio and PyCharm). 

First, install R from [https://www.r-project.org/](https://www.r-project.org/). Note that you will have to choose a CRAN mirror that you wish to download R from - which one you pick does not matter. Since I am based in Singapore I typically choose one from nearby regions (Taiwan, Korea, Japan, Indonesia, or Worldwide). Download the latest version. 

Then, install RStudio from [https://posit.co/](https://posit.co/). Download the correct RStudio Desktop for your computer OS. ([https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/))

Note that both R and RStudio are free downloads. You should not need to pay for anything. 

## RStudio Cloud

If you do not have a computer/laptop that you are able to install programs on, you can try using [RStudio in the cloud](https://posit.cloud/). Basically, you can run RStudio on the internet without downloading anything into your computer. 

The main consideration is that the free version has a cap on usage. Below is reproduced from their website ([https://posit.cloud/plans/free](https://posit.cloud/plans/free)):

> "Compute hours represent how much time a project is open or running with a particular configuration, computed as (RAM + CPUs allocated) / 2 x hours. For example, working with a project for 1 hour with 1 GB of RAM and 1 CPU allocated consumes 1 project hour. The Cloud Free plan has a cap each month of 25 project hours. Once you reach the cap, you can no longer open or create projects during your current month."

## Installing R packages 

Now that you have R and RStudio set up, let's try to install the R packages that we will use in this book. First, what is a "R package"? Sometimes they are known as libraries, but you can think of packages/libraries as a collection of R functions that do specialized things. Because we make use of several functions from the `igraph` package to do network analysis, we need to first download `igraph`. 

To download `igraph`, open RStudio and type the following into your Console (the panel with the ">" symbol - this is where you type in instructions for R to execute). 

```{r eval=FALSE}
install.packages("igraph")
```

Be patient and wait for the package to finish downloading. After it is downloaded, we can only start using the functions in `igraph` *after* we have loaded the library into our workspace. To do this, type the following into the Console: 

```{r eval=TRUE}
library(igraph)
```

If everything is working fine, you should see a plot of two connected nodes when you run the following line of code in the Console: 

```{r eval=TRUE}
plot(graph_from_literal(A-B))
```

**Important!** You only need to install a package **ONCE** (unless you wish to update it). But you will always need to load the package at the start of the R session (i.e., run `library(igraph)` in the Console whenever you open RStudio) so that you can use these special network functions. 

### Exercise

1. Try installing the package called `tidyverse`. `tidyverse` is a powerful set of packages that enable people to do cool data science things in R. It is highly recommended to learn it if you are interested in becoming a proficient user of R. See [https://www.tidyverse.org/](https://www.tidyverse.org/) for more information.

2. You know if you have succeeded if you are able to load `tidyverse` into your workspace. What is the code needed to do this? 
