---
output:
  pdf_document: default
  html_document: default
---
# Chapter 12: `influenceR` package {#ch12}

In this chapter, we explore two additional micro-level network metrics that are developed in the domain of social network analysis. For this, we need to install the `influenceR` R package which provides the functions for computing bridging scores and effective network sizes for nodes in the network.

## Set up 

First, let's install the `influenceR` R package:

```{r}
#| echo: true
#| eval: false
install.packages('influenceR')
```

We will continue to use the `macaque` network from Chapter 11 for demonstration. 

```{r}
#| echo: true
#| eval: true
library(igraphdata)

data("macaque")

# load the other packages that we need 
library(igraph)
library(influenceR)
```

## Bridging score

Granovetter (1973) observed that edges that reduce the overall distance of a network are structurally important bridges. These are usually the "weak" edges in a social network where the nature of the relationship between those two individuals are on an acquaintance basis. Nevertheless, the presence of these edges enable connection between distant parts of the network. 

Valente and Fujimoto (2010) built on that insight and developed a "node-centric" measure of bridging that considers the impact of deleting an edge or link on the cohesiveness of the network. In other words, a node with a higher bridging score tends to possess a structurally important edge, such that deleting it substantially reduces the overall connectivity of the network. 

A node’s bridging score, $B_i$, is the average decrease in cohesiveness when each of its edges are systematically removed from the graph.

$B_i=\frac{∑ΔC}{k_i}$

where $k_i$ = the degree of node *i* and $ΔC$ = change in network cohesiveness.

For each connected edge of a target node:

1. Remove the edge and recompute network cohesiveness (inverse of distance)
2. Take the difference from the original network cohesiveness (i.e., the decrease)

Then, take the sum of the change in cohesiveness, divide by the number of edges (degree).

Below, we can easily use the `bridging` function to compute this information for us. 

```{r}
#| echo: true
#| eval: true
bridging(macaque) # from influenceR

bridging_data <- data.frame(node = V(macaque)$name,
                            bridge_score = bridging(macaque))

head(bridging_data)
```

## Effective network science 

Burt (2004) argues that people who are located near "holes" in a social network tend to have better ideas. This occurs because such individuals are "brokers" across different social groups and can accumulate social capital more quickly.

The effective size of a node's ego network is based on the concept of redundancy. A person's ego network has redundancy to the extent that their contacts are connected to each other as well. The nonredundant part of a person's relationships is the effective size of their ego network. The maximum ENS is the degree of the ego. The minimum ENS is 1, as if one "effectively" only had 1 single friend.

A node’s effective network size, $E_i$, is computed as follows:

$E_i=k−2t/k$

where *t* = number of ties in the ego network and *k* = the degree of the node.

For each node: 

1. Extract the ego network of the node. The ego network is the immediate neighborhood of the node.
2. Check the extent to which its neighbors are connected by counting number of ties, *t*.

Below, we can easily use the `ens` function to compute this information for us. 

```{r}
#| echo: true
#| eval: true
ens(macaque) # from influenceR

ens_data <- data.frame(node = V(macaque)$name,
                            ens_score = ens(macaque))

head(ens_data)
```

By now, you should know how to export the data frames as a .csv file for further exploration! (Hint: `write.csv(...)`).

## References 

Burt, R. S. (2004). Structural holes and good ideas. *American Journal of Sociology, 110*(2), 349-399.

Granovetter, M. S. (1973). The strength of weak ties. *American Journal of Sociology, 78*(6), 1360-1380.

Valente, T. W., & Fujimoto, K. (2010). Bridging: locating critical connectors in a network. *Social Networks, 32*(3), 212-220.

## Exercise 

Conduct the following analysis on a social network of your choosing, where the nodes correspond to individual persons. If you do not have a social network, you could conduct the analysis on the `karate` or `UKfaculty` networks from the `igraphdata` library.

1. Identify the individuals with the highest and lowest bridging scores.

- What is the implication of having a high bridging score for this particular network?
- What is the implication of having a low bridging score for this particular network?

2. Identify the individuals with the highest and lowest effective network sizes.

- What is the implication of having a high effective network size for this particular network?
- What is the implication of having a low effective network size for this particular network?
